Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-25 paper/main.tex
more slack discussion-inspired changes-- updating description of putting stop word list together

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. We note that we used these 2D maps solely for
visualization; all relevant comparisons, distance computations, and statistical
tests we report above were carried out in the original 15-dimensional space,
using the 15-topic model. Aside from increasing the number of topics from 15 to
100, all other procedures and model parameters were carried over from the
preceding analyses. As in our other analyses, we resampled each lecture's topic
trajectory to 1~Hz and projected each question into a shared text embedding
space.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model~\citep[Latent
Dirichlet Allocation; ][]{BleiEtal03} trained on a set of documents, to
discover a set of (up to) $k$ ``topics'' or ``themes.'' Formally, each topic is
defined as a distribution of weights over words in the model's vocabulary
(i.e., the union of all unique words, across all documents, excluding ``stop
words.''). Conceptually, each topic is intended to give larger weights to words
that are semantically related (as inferred from their tendency to co-occur in
the same document). After fitting a topic model, each document in the training
set, or any \textit{new} document that contains at least some of the words in
the model's vocabulary, may be represented as a $k$-dimensional vector
describing how much the document (most probably) reflects each topic. To select
an appropriate $k$ for our model, as a starting point, we identified the
minimum number of topics that yielded at least one ``unused'' topic (i.e., in
which all words in the vocabulary were assigned uniform weights) after
training. This indicated that the number of topics was sufficient to capture
the set of latent themes present in the two lectures (from which we constructed
our document corpus, as described below). We found this value to be $k = 15$
topics. We found that with a limited number of additional adjustments
following~\citep{BoydMimn14}, such as removing corpus-specific stop-words, the
model yielded (subjectively) sensible and coherent topics. The distribution of
weights over words in the vocabulary for each discovered topic is shown in
Supplementary Figure~\topicWordWeights, and each topic's top-weighted words may
be found in Supplementary Table~\topics.

>>>>

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. We note that we used these 2D maps solely for
visualization; all relevant comparisons, distance computations, and statistical
tests we report above were carried out in the original 15-dimensional space,
using the 15-topic model. Aside from increasing the number of topics from 15 to
100, all other procedures and model parameters were carried over from the
preceding analyses. As in our other analyses, we resampled each lecture's topic
trajectory to 1~Hz and projected each question into a shared text embedding
space.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model~\citep[Latent
Dirichlet Allocation; ][]{BleiEtal03} trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is
defined as a distribution of weights over words in the model's vocabulary
(i.e., the union of all unique words, across all documents, excluding ``stop
words.''). Conceptually, each topic is intended to give larger weights to words
that are semantically related (as inferred from their tendency to co-occur in
the same document). After fitting a topic model, each document in the training
set, or any \textit{new} document that contains at least some of the words in
the model's vocabulary, may be represented as a $k$-dimensional vector
describing how much the document (most probably) reflects each topic. To select
an appropriate $k$ for our model, as a starting point, we identified the
minimum number of topics that yielded at least one ``unused'' topic (i.e., in
which all words in the vocabulary were assigned uniform weights) after
training. This indicated that the number of topics was sufficient to capture
the set of latent themes present in the two lectures (from which we constructed
our document corpus, as described below). We found this value to be $k = 15$
topics. We found that with a limited number of additional adjustments
following~\citep{BoydMimn14}, such as removing corpus-specific stop-words, the
model yielded (subjectively) sensible and coherent topics. The distribution of
weights over words in the vocabulary for each discovered topic is shown in
Supplementary Figure~\topicWordWeights, and each topic's top-weighted words may
be found in Supplementary Table~\topics.
